Chasing \rainbows{} and ocean glint:\\ Inner working angle constraints for the Habitable Worlds Observatory

### 2023-05-01 src/tex/ms.tex
[showyourwork] overleaf sync

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\hwo), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight that is reflected by an exoplanet can reveal angular features that are due to ocean glint, and to rainbows and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \hwo's coronagraph may obscure the phase angles where such characteristic features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
Planets that orbit close to edge-on have accessible scattering angles limited by coronagraph obscuration. 
%
We use \hwo'~catalog of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering by gas. 
%
We find that any exo-Earths in systems in this catalog will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The glint signature at planetary scattering angles of $\sim\qtyrange{50}{70}{\degree}$ 
%\sim$110--130$^\circ$ 
would be accessible to \hwo~in $\sim\num{14}$ systems, and the rainbow due to water clouds at phase angles of \qtyrange{140}{160}{\degree} would be accessible in $\sim\num{42}$ systems, assuming a \qty{62}{\mas} \IWA{} ($3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Increasing the \IWA{} to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with the planet's orbital eccentricity and scales approximately inversely with the \IWA{}.  \end{abstract}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/Fig1IWAPol.png}
    \caption{
        The phase curves of two homogeneous planets with oceans and Earth-like atmospheres in edge-on orbits (the planets thus exhibit scattering angles 
        ranging from 0$^\circ$ to 180$^\circ$ along their orbits). 
        The cloudy scenario includes a low-altitude, Earth-like water 
        cloud deck. 
        Below the plot, we are showing the scattering angle ranges that cover the
        characteristic features of glories, rainbows, Rayleigh scattering, glint, and forward scattering by cloud particles (note that features partly overlap
        with each other).  
        Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
    }
    \label{fig:bottplot}
\end{figure*}

@@

The field of exoplanet research has made great advances in the last few decades. The characterisation of Earth-like planets in the habitable zones of their stars is now at the frontier of what we can achieve. The Habitable Worlds Observatory will perform a dedicated direct imaging survey of nearby stars searching for rocky planets in the habitable zone. This space-based coronographic telescope will detect and characterise exoplanets using their reflection spectrum. Complementary information on the habitability of these worlds can be obtained by observing the change in the planet's brightness at different planetary phases.

@@

An exoplanet observed in an edge-on orbit will move through all phases and thus all phase angles, while those viewed at higher inclinations will pass through a subset of these phase angles.
%
If the planet is rotating, its reflected light signal in both total and polarized flux is expected to show temporal variations that can place constraints on the horizontal heterogeneity of the planet \citep{2001Natur.412..885F,stam2008,2022A&A...664A..59M}.

The strengths of detailed total and polarized flux phase curve analysis include mapping \citep{2001Natur.412..885F, berdyugina2019surface, 2022A&A...664A..59M}, the detection of dark areas or glint indicative of seas \citep{groot2020, cowan2008inverting, lustig2019}, and the characterization of clouds through forward- and back-scattering, rainbows, and glories.
%
An added potential benefit of observing the state of polarization of the reflected light is the suppression of starlight from quiescent stars without explicitly requiring a coronagraph or starshade as the light of these stars themselves can be considered to be unpolarized \citep{kemp1987}.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces. While glint is a familiar phenomenon on Earth, there is only
one other body in the Solar System where it has been observed: 
the Cassini orbiter captured the glint of sunlight reflecting off the
liquid hydrocarbon lakes near Titan's north pole as it was observing
the moon from the nightside \citep{2010GeoRL..37.7104S}.
%
In the total flux, the glint signal is indeed expected to be most prominent at large planetary phase angles.
%
The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at \qty{53}{\degree} \citep[i.e., a phase angle of \qty{127}{\degree}; see, e.g.,][]{2008Icar..195..927W}.
%
This angle therefore varies slightly with the wavelength and the composition of the ocean and the atmosphere.
%

In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The stronger glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
The width across phase angles of the glint feature in total and polarized flux generally
broadens with increasing wave height \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}, which depends strongly on the wind speed over the surface, as described by \citep{CoxMunk1954}.
%
{\color{red}{check the next sentence:}}
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across phase angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

The glint provides evidence of liquid surface bodies, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
At large phase angles, the forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which further motivated probing the rainbow angle for context.
%
This confusion can be solved by measuring the intersection of phase curves at various wavelengths in polarized light (the polarized colour changes from blue, through white, to red), which only occurs when the glint contributes to the signal and at a planetary scattering angle depending on the cloud coverage fraction \citep{treesandstam2019}.
%The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
%angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
While the glint feature can only be probed when a planet is at crescent phases, \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry when the planet is at scattering angles where it is at the wide separations more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

Rainbows, cloudbows, glories and other optical phenomena from light that is 
scattered by atmospheric condensates can dramatically influence the shape of a planet's total flux and polarized phase curves and provide a wealth of information 
about exoplanets \citep{karalidi2012rainbow, stam2008,Bailey2007,2014A&A...566L...1G}.%
\footnote{
    Rainbows occur when light is scattered by rain droplets that are large compared to the wavelength of the light. Cloudbows refer to features occurring at the same phase angles from smaller cloud particles with the spread in colors invisible to the human eye making them appear white \citet[see][for further discussion]{Bailey2007}.  
}
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, the polarized phase curve of Venus at three wavelengths showed not only that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}. 
The glory feature in Venus's total flux phase curve was later used to rule out potential solutions for the unknown UV absorber \citep{petrova2018glory}.

In unpolarized light, the scattering by cloud particles can dramatically change the shape of a planet's phase curve compared to that of a 
Lambertian reflecting planet.
%
While the glory and rainbow are technically visible both in the total flux
and the polarized flux phase curves
in the presence of spherical cloud and/or haze particles, their strongly polarized nature coupled with the suppression of signals from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized flux than in total flux \citep{karalidi2011, stam2008, treesandstam2019}.
%
The rainbow angle depends on the refractive index of the scatterers and thus depends on their composition and the wavelength. 
The wavelength dependence itself also depends on the particles' size distribution, and the size distribution affects the angular shape of the rainbow \citep{karalidi2011}
%
Because the rainbow feature is narrower than the glint due to the surface reflection, it would 
allow for a more exact determination of the composition of the condensates.
%
Note that the scattering by clouds that consist of solid particles, 
such as ice crystals, will produce other optical features, such as halo's, particularly at small phase angles \citep{bailey2007, hansentravis1974, karalidi2012rainbow}.
%

In this paper, we define our optimal angle for the detection of a rainbow as that of the peak of the water droplet rainbow at visible wavelengths, which is around a phase angle $\alpha$ of 40$^\circ$ (see Fig.~\ref{fig:bottplot}).
%
Pushing observations to angles closer to the star (smaller phase angles, i.e., larger scattering angles) would allow the determination of the full shape of this rainbow, and enable the identification of rainbows due to the scattering of light by other possible HZ condensates.
%
The glory 
%is produced when the droplet radii much smaller than the wavelength of light 
can allow further characterization of condensates as its angle and shape also depend on the complex refractive index, but this feature occurs within \qty{20}{\degree} phase angle very close to full phase \citep{hansentravis1974}.
%
%The glory (in the backscattering direction) and diffraction peak (in the forward scattering direction) are distinct features in the total reflected flux, but not in the degree of polarization. In principle, the degree of polarization goes to zero in the forward and backward scattering directions. See e.g. Fig. 1 of \citet{stam2008}, Fig. 5 of \citet{trees2022}, Fig. 3 of \citet{karalidi2011} and \citet{hansentravis1974}.}}

@@

In this paper, we quantify the planetary scattering angles that would be accessible for direct observations for hypothetical terrestrial planets orbiting in the habitable zones of \hwo{} target stars.
%
In \cref{subsec:2.1}, we describe the target list, in \cref{subsec:2.2} we derive expressions for the range of scattering angles given a few different limits, in \cref{subsec:2.3}, we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets in these systems, and in \cref{subsec:2.4}, we outline how the planet/star contrast changes as a function of the planet's orbital phase.  
%
We present and discuss our results in \cref{sec:3}, and conclude in \cref{sec:4}. 
It is important to note that the phases angles selected for the features throughout this paper are approximate, that features overlap, and that their maximum peak will usually vary with chemical species and wavelength.

The conventions for angle names vary throughout the literature on exoplanet phase curves. In this paper we define the \textbf{scattering} angle as the local divergence in the light direction (large angles for back-scattering, small angles for forward-scattering); the \textbf{phase} angle $\varphi$ as the viewer-star-planet angle, which is zero at full phase and 180 at new phase, and which is the complement of the scattering angle. A planet with a face-on orbit will thus have a phase angle value of 90 degrees for the duration of its orbit.  In some plots we ``fold'' the half orbit so that we measure \textbf{degrees from quadrature} (so that both new and full phase are at maxima of 90$^\circ$ for edge on orbits), which is helpful in considering the visibility of the planet with inner working angle. We also note that we are considering here the polarization flux which takes into account the intensity of the polarization and reflectance (illumination) of the planet but additional information can be garnered from full Stokes characterization.

@@

%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo{} is still in the early stages of development, and although the exact telescope and instrument design have not yet been determined, we assume a \qty{6}{\meter} primary mirror, as suggested in The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey.
%
The observatory will utilise a coronagraph to suppress the stellar light and facilitate high-contrast observations of faint accompanying exoplanets. 
%
The \IWA{} of the coronagraph will be driven by both the science requirements and technological limitations and is still unknown.
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA{} of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
%From the HabWorlds report. Inner Working Angle: The IWA defines the region near the star that cannot be accessed for direct imaging due to a coronagraphic mask, starshade obscuration, or an interferometric null. It depends on the architecture of the telescope, the starlight suppression system used, and the observation wavelength. While a nominal value could be defined by making assumptions about the design for Habitable Worlds Observatory, we choose instead to derive the IWA from the star list itself and Astro2020’s requirement that ~100 cumulative habitable zones be surveyed by direct imaging. As shown in the main table, an IWA near ~70 mas will be needed to access ~100 cumulative habitable zones, and would need to be achieved at all wavelengths of interest for spectral characterization. For any specific telescope and starlight suppression system architecture the IWA is typically proportional to wavelength; thus, the number of accessible HZs will strongly decrease as IWA(lambda) increases.
For ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a \qty{6}{\meter} telescope at representative wavelengths $\lambda$ of $\qty{600}{\nano\meter}$ and $\qty{1}{\micro\meter}$.

@@

In what follows, we consider hypothetical Earth-sized planets orbiting at the right distance from their parent star to receive the same flux as the Earth receives from the Sun.
%
By design, such planets should be observable by \hwo{} at least at their quadrature phase ($\alpha=90^\circ$).

@@

In the non-eccentric case (i.e., a circular orbit with the coronagraphic mask centred on the star), the minimum ($\varphi_\mathrm{min}$) and maximum ($\varphi_\mathrm{max}$) accessible phase angles are symmetric about 90$^\circ$. 
%
We define the angle $\Delta \varphi$ such that: 
\begin{equation}
 \label{eq:Delta_phi}
    \Delta \varphi 
    = \varphi_\mathrm{max} - \varphi_\mathrm{min}
    =  2 \cdot (\varphi_\mathrm{max} - \qty{90}{\degree}) 
    =  2 \cdot (\qty{90}{\degree} - \varphi_\mathrm{min} )  \,.
\end{equation} 
The on-sky projected planet distance, $r_\mathrm{proj}$ (in \unit{\au}) is shown in \cref{fig:scattering-angle} for a non-eccentric case with a 
semi-major orbital axis $a$ (in \unit{\au}), at a distance $d_*$ (in \unit{\parsec}) from the observer. 
%
The phase angle $\varphi$ is given by
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a} \,.
    \label{eq:sinphi}
\end{equation}
%
Assuming small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the on-sky angular separation.
%
We access the minimum phase angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Combining \cref{eq:Delta_phi} and \cref{eq:sinphi}, we deduce:
%
\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA} 
    \cdot d_*}{a} \,.
\end{equation}

@@

\Cref{fig:scatterplot} shows $\Delta \varphi$ for the systems in the \hwo{} target list, assuming an IWA of \qty{62}{\mas} ($3\,\lambda / D$ coronagraph at $\qty{600}{\nano\meter}$, see \cref{tab:IWA_OWA}). 
%
%This assumes each planet has a semi-major axis such that it receives an Earth-equivalent instellation, which is provided in the input star list.
%
As can be seen in \cref{fig:scatterplot}, for exoplanets in edge-on circular orbits, it will always be possible to observe the angular feature that is representative for Rayleigh scattering, as that occurs around quadrature ($\alpha=90^\circ$), while the rainbow angle (around $\alpha=40^\circ$) will only be accessible for a subset of exoplanets.

%--------------------------------------------------------------------------
% Figure 4
%--------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics{figures/figure-4-scatterplot.pdf}
    \script{create-scatterplot.py}
    \caption{
        Scatter plot for the target star sample, showing stellar effective 
        temperatures and distances.
        %
        The circle sizes represent the angular separation between 
        the star and its planet (in mas) as presented in the target list.
        %
        The colours show $\Delta \varphi$ assuming circular, edge-on orbits
        at a semi-major axis $a$ corresponding to an Earth-like 
        instellation and an IWA of \qty{62}{\mas}.
        %
        Additionally, the colour bar indicates the optical phenomena that 
        can in principle be detected: the phenomena from the bottom of the bar up to the colour of a given circle would be detectable for that planet.
        %
        For example, dark blue circles indicate planets on which most phenomena would be observable because planets that can be observed at the rainbow angle can also be observed at angles exhibiting Rayleigh scattering.
    }
    \label{fig:scatterplot}
\end{figure*}
%--------------------------------------------------------------------------

>>>>

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\hwo), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight that is reflected by an exoplanet can reveal angular features that are due to ocean glint, and to rainbows and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \hwo's coronagraph may obscure the phase angles where such characteristic features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
Planets in close-to-edge-on orbits have accessible phase angles limited by coronagraphic obscuration. 
%
We use \hwo'~catalog of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering by gas. 
%
We find that any exo-Earths in systems in this catalog will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The ocean glint signature at planetary scattering angles of $\sim\qtyrange{50}{70}{\degree}$ 
%\sim$110--130$^\circ$ 
would be accessible to \hwo~in $\sim\num{14}$ systems, and the rainbow due to water clouds at phase angles of \qtyrange{140}{160}{\degree} would be accessible in $\sim\num{42}$ systems, assuming a \qty{62}{\mas} \IWA{} ($3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Increasing the \IWA{} to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with the planet's orbital eccentricity and scales approximately inversely with the \IWA{}.  \end{abstract}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/Phase_curves.pdf}
    \caption{The phase curves of the polarized flux of a planet with an ocean surface and an Earth-like atmosphere in an edge-on orbit (the planet thus exhibits phase angles ranging from 0$^\circ$ to 180$^\circ$ along its orbit). The cloud fraction is 0.5, the cloud pattern is patchy, and the glint is kept cloud-free to maximize the glint signature \citep[see ][]{treesandstam2019}.
        %The cloudy scenario includes a low-altitude, Earth-like water 
        %cloud deck. 
        Below the plot, we are showing the phase angle ranges that cover the
        characteristic features of glories, rainbows, Rayleigh scattering, glint, and forward scattering by cloud particles (note that features partly overlap
        with each other).  
        Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
    }
    \label{fig:bottplot}
\end{figure}

@@

@@

An exoplanet in an edge-on orbit will move through phase angles ranging from $0^\circ$ (full phase), through $90^\circ$ (quadrature), and to $180^\circ$ (the planet's nightside is towards the observer), while planets higher inclination orbits will pass through a subset of phase angles.
%
If the planet is rotating, its reflected light signal is expected to show temporal variations that can place constraints on the horizontal heterogeneity of the planet.

The strengths of detailed phase curve analysis include mapping \citep{2001Natur.412..885F, berdyugina2019surface, 2022A&A...664A..59M}, the detection of dark areas or glint indicative of seas \citep{groot2020, cowan2008inverting, lustig2019}, and the characterization of clouds through forward- and back-scattering, rainbows, and glories.
%
An added potential benefit of observing the state of polarization of the reflected light is the suppression of starlight from quiescent stars without explicitly requiring a coronagraph or starshade as the light of these stars themselves can be considered to be unpolarized \citep{kemp1987}.

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces. While glint is a familiar phenomenon on Earth, there is only
one other body in the Solar System where it has been observed: 
the Cassini orbiter captured the glint of sunlight reflecting off the
liquid hydrocarbon lakes near Titan's north pole as it was observing
the moon from the nightside \citep{2010GeoRL..37.7104S}.
%
In the total flux, the glint signal is indeed most prominent at small planetary phase (large scattering) angles.
%
The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at \qty{53}{\degree} \citep[i.e., a phase angle of \qty{127}{\degree}; see, e.g.,][]{2008Icar..195..927W}.
%
This angle therefore varies slightly with wavelength and with sea and atmosphere composition.
%
In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The strengthened glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
{\color{red}{check the next sentence:}}
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across phase angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

The glint provides evidence of liquid surface bodies, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
The forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which further motivated probing the rainbow angle for context.
%
This confusion can be solved by measuring the intersection of phase curves at various wavelengths in polarized light (the polarized colour changes from blue, through white, to red), which only occurs when the glint contributes to the signal and at a planetary scattering angle depending on the cloud coverage fraction \citep{treesandstam2019}.
%The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
%angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
While the glint feature can only be probed when a planet is at crescent phases, \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry when the planet is at scattering angles where it is at the wide separations more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

Rainbows, cloudbows, glories and other optical phenomena from light that is 
scattered by atmospheric condensates provide a wealth of information 
about exoplanets \citep{karalidi2012rainbow, stam2008,2014A&A...566L...1G}.%
\footnote{
    Rainbows occur when light is scattered by rain droplets, which are large compared to the wavelength of the light. Cloudbows refer to features occurring at the same phase angles from smaller cloud particles with the spread in colors invisible to the human eye making them appear white (\citet[see][for further discussion]{Bailey2007}).  
}
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, the phase curve behaviour of Venus showed not only that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}. 
The glory feature in Venus's phase curve was later used to rule out some solutions for the unknown UV absorber \citep{petrova2018glory}.

In unpolarized light, the scattering by cloud particles can dramatically change the shape of a planet's phase curve compared to that of a 
Lambertian reflecting planet.
%
{\color{red}{check the sentence:}}
While the glory and rainbow are technically visible in any phase curve
in the presence of spherical cloud and/or haze particles, their strongly polarized nature coupled with the suppression of signals from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized light (both in the degree of polarization and in full Stokes measurements) \citep{karalidi2011, stam2008, treesandstam2019}.
%
The rainbow angle depends on the refractive index of the scatterers and thus depends on the composition and the wavelength. 
The wavelength dependence itself also depends on the particles' size distribution, and the size distribution affects the angular shape of the rainbow \citep{karalidi2011}
%
Because the rainbow feature is more narrow than the glint, it would 
allow for a more exact determination of the particle composition.
%
Note that the scattering by clouds that consist of solid particles, 
such as ice crystals, will produce other optical features, such as halo's, particularly at small phase angles \citep{bailey2007, hansentravis1974, karalidi2012rainbow}.
%

In this paper, we define our optimal angle for the detection of a rainbow as that of the peak of the water droplet rainbow at visible wavelengths.
%
Pushing observations to angles closer to the star (smaller phase angles, i.e., larger scattering angles) would allow the determination of the full shape of this rainbow, and enable the identification of rainbows due to the scattering of light by other possible HZ condensates.
%
The glory 
%is produced when the droplet radii much smaller than the wavelength of light 
can allow further characterization of condensates as its angle and shape also depend on the complex refractive index, but this feature occurs within \qty{20}{\degree} phase angle very close to full phase \citep{hansentravis1974}.
%
%The glory (in the backscattering direction) and diffraction peak (in the forward scattering direction) are distinct features in the total reflected flux, but not in the degree of polarization. In principle, the degree of polarization goes to zero in the forward and backward scattering directions. See e.g. Fig. 1 of \citet{stam2008}, Fig. 5 of \citet{trees2022}, Fig. 3 of \citet{karalidi2011} and \citet{hansentravis1974}.}}

@@

In this paper, we quantify the planetary phase angles that would be accessible for direct observations for hypothetical terrestrial planets orbiting in the habitable zones of \hwo{} target stars.
%
In \cref{subsec:2.1}, we describe the target list, in \cref{subsec:2.2} we derive expressions for the range of phase angles given a few different limits, in \cref{subsec:2.3}, we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets in these systems, and in \cref{subsec:2.4}, we outline how the planet/star contrast changes as a function of the planet's orbital phase.  
%
We present and discuss our results in \cref{sec:3}, and conclude in \cref{sec:4}. 
It is important to note that the phases angles selected for the optical phenomena throughout this paper are approximate, that phenomena overlap, and that their maximum peak will usually depend on the chemical species and wavelength.

The conventions for angle names vary throughout the literature on exoplanet phase curves. In this paper we define the \textbf{scattering} angle as the local divergence in the propagation direction of the light (large angles for back-scattering, small angles for forward-scattering); the \textbf{phase} angle $\varphi$ as the viewer-star-planet angle, which is zero at full phase and 180 at new phase, and which is the complement of the scattering angle. A planet with a face-on orbit will thus have a phase angle value of 90 degrees for the duration of its orbit. In some plots we ``fold'' the half orbit so that we measure \textbf{degrees from quadrature} (so that both new and full phase are at maxima of 90$^\circ$ for edge on orbits), which is helpful in considering the visibility of the planet with inner working angle. We also note that we are considering here the polarization flux which takes into account the intensity of the polarization and reflectance (illumination) of the planet but additional information can be garnered from full Stokes characterization.

@@

%\textcolor{blue}{Merge the 2 sections here into "Assumed technical specifications and sample"? I don't think this needs to be 2 subsections}
\hwo{} is still in the early stages of development, and although the exact telescope and instrument design have not yet been determined, we assume a \qty{6}{\meter} primary mirror, as suggested in The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey.
%
The observatory will utilise a coronagraph to suppress the stellar light and facilitate high-contrast observations of faint accompanying exoplanets. 
%
The \IWA{} of the coronagraph will be driven by both the science requirements and technological limitations and is still unknown.
%In this work, we investigate the scattering phase coverage obtainable for several choices for the \IWA{} of the observatory. 
%We also consider the Outer Working Angle (OWA) but for most targets this is not a limiting factor. 
%From the HabWorlds report. Inner Working Angle: The IWA defines the region near the star that cannot be accessed for direct imaging due to a coronagraphic mask, starshade obscuration, or an interferometric null. It depends on the architecture of the telescope, the starlight suppression system used, and the observation wavelength. While a nominal value could be defined by making assumptions about the design for Habitable Worlds Observatory, we choose instead to derive the IWA from the star list itself and Astro2020’s requirement that ~100 cumulative habitable zones be surveyed by direct imaging. As shown in the main table, an IWA near ~70 mas will be needed to access ~100 cumulative habitable zones, and would need to be achieved at all wavelengths of interest for spectral characterization. For any specific telescope and starlight suppression system architecture the IWA is typically proportional to wavelength; thus, the number of accessible HZs will strongly decrease as IWA(lambda) increases.
For ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a \qty{6}{\meter} telescope at representative wavelengths of $\lambda = \qty{600}{\nano\meter}$ and $\qty{1}{\micro\meter}$.

@@

In what follows, we consider hypothetical Earth-sized planets orbiting at the right distance from their parent star to receive the same flux as the Earth receives from the Sun.
%
By design, such planets should be observable by \hwo{} at least at their quadrature phase.

@@

In the non-eccentric case (i.e., a circular orbit with the coronagraphic mask centred on the star), the minimum ($\varphi_\mathrm{min}$) and maximum ($\varphi_\mathrm{max}$) accessible phase angles are symmetric about 90$^\circ$. 
%
We define the angle $\Delta \varphi$ such that: 
\begin{equation}
 \label{eq:Delta_phi}
    \Delta \varphi 
    = \varphi_\mathrm{max} - \varphi_\mathrm{min}
    =  2 \cdot (\varphi_\mathrm{max} - \qty{90}{\degree}) 
    =  2 \cdot (\qty{90}{\degree} - \varphi_\mathrm{min} )  \,.
\end{equation} 
The on-sky projected planet distance, $r_\mathrm{proj}$ (in \unit{\au}) is shown in \cref{fig:scattering-angle} for a non-eccentric case with a 
semi-major orbital axis $a$ (in \unit{\au}), at a distance $d_*$ (in \unit{\parsec}) from the observer. 
%
The phase angle $\varphi$ is given by
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a} \,.
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum phase angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:
%
\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA} 
    \cdot d_*}{a} \,.
\end{equation}

@@

\Cref{fig:scatterplot} shows $\Delta \varphi$ for the systems in the \hwo{} target list, assuming an IWA of \qty{62}{\mas} ($3\,\lambda / D$ coronagraph at $\qty{600}{\nano\meter}$, see \cref{tab:IWA_OWA}). 
%
%This assumes each planet has a semi-major axis such that it receives an Earth-equivalent instellation, which is provided in the input star list.
%
As can be seen in \cref{fig:scatterplot}, for exoplanets in edge-on circular orbits, it will always be possible to observe the angular feature that is representative for Rayleigh scattering, as that occurs around quadrature, while the rainbow angle will only be accessible for a subset of exoplanets.

%--------------------------------------------------------------------------
% Figure 4
%--------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics{figures/figure-4-scatterplot.pdf}
    \script{create-figure-4-scatterplot.py}
    \caption{
        Scatter plot for the target star sample, showing stellar effective 
        temperatures and distances.
        %
        The circle sizes represent the angular separation between 
        the star and its planet (in mas) as presented in the target list.
        %
        The colours show $\Delta \varphi$ assuming circular, edge-on orbits
        at a semi-major axis $a$ corresponding to an Earth-like 
        instellation and an IWA of \qty{62}{\mas}.
        %
        Additionally, the colour bar indicates the optical phenomena that 
        can in principle be detected: the phenomena from the bottom of the bar up to the colour of a given circle would be detectable for that planet.
        %
        For example, dark blue circles indicate planets on which most phenomena would be observable because planets that can be observed at the rainbow angle can also be observed at angles exhibiting Rayleigh scattering.
    }
    \label{fig:scatterplot}
\end{figure*}
%--------------------------------------------------------------------------
